---
title: "3 - How to use the lifelihood package"
author: "Nicolas RODE, Thomas LENORMAND"
date: "`r format(Sys.Date(), '%d-%B-%Y')`"
output: 
  html_document:
    theme: journal
    css: style.css
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

<br>

If you haven't already check it, have a look at:

- [what is the required data format to work with lifelihood?](1-required-data-format.html)
- [setting up the configuration file](2-setting-up-the-configuration-file.html)

# Load libraries
***

```{r}
library(lifelihood)
```

<br>

# Data preparation
***

Load the dataset from `.csv` file:

```{r}
# input data
df <- read.csv(here::here('data/fake_sample.csv'))
```

```{r, echo=FALSE}
DT::datatable(df, options = list(scrollX = TRUE), class = "cell-border stripe hover compact")
```

Prepare input parameters for the `lifelihood()` function:

```{r}
# name of the columns of the clutchs into a single vector
clutchs <- c(
  "clutch_start1", "clutch_end1", "clutch_size1",
  "clutch_start2", "clutch_end2", "clutch_size2"
)

# seeds value (must of length 4!)
seeds <- c(1, 2, 3, 4)
```

<br>

# Get the results
***

## All default parameters

```{r}
# run the analysis
results <- lifelihood(
  df = df,
  path_config = "../config.yaml",
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "mor_start",
  death_end = "mor_end",
  covariates = c("geno", "type"),
  seeds = seeds,
  matclutch = FALSE,
  models = c("gam", "lgn", "wei")
)
summary(results)
```

## With custom parameter ranges

```{r}
# dataframe with param ranges
param_ranges <- data.frame(
  param = c("E(tmort)f", "morta", "RE(tmort)m", "mortp", "propmal", "E(tmat)f", "mata", 
            "RE(tmat)m", "E(tpon)", "ponta", "pontn", "to(ps)int", "to(ps)am", "to(ps)tp", 
            "sen(pu)t", "sen(pu)t2", "sen(pn)t", "sen(pn)t2", "to(pupn)", "W"),
  min = c(1, 0.001, 0.1, 0.0001, 0.00001, 1, 0.0001, 0.1, 0.1, 0.001, 1, 0.00001, 
          0.0000001, 0.0000001, -20, -20, -8, -8, -8, 0.001),
  max = c(301, 20, 4, 1, 0.99999, 100, 15, 10, 200, 12, 50, 10, 10, 10, 20, 20, 10, 10, 10, 1200)
)
```

```{r, echo=FALSE}
DT::datatable(param_ranges, options = list(scrollX = TRUE), class = "cell-border stripe hover compact")

# run the analysis
results <- lifelihood(
  df = df,
  path_config = "../config.yaml",
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "mor_start",
  death_end = "mor_end",
  covariates = c("geno", "type"),
  seeds = seeds,
  param_range_df = param_ranges,
  matclutch = FALSE,
  models = c("gam", "lgn", "wei")
)
summary(results)
```

<br>


# Get specific results
***

```{r}
results$effects$Estimation1
```