---
title: "3 - How to use the lifelihood package"
author: "Nicolas RODE, Thomas LENORMAND"
date: "`r format(Sys.Date(), '%d-%B-%Y')`"
output: 
  html_document:
    theme: journal
    css: style.css
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

<br>

If you haven't already check it, have a look at:

- [what is the required data format to work with lifelihood?](1-required-data-format.html)
- [setting up the configuration file](2-setting-up-the-configuration-file.html)

# Load libraries
***

```{r}
#library(lifelihood)
devtools::load_all()
```

<br>

# Data preparation
***

Load the dataset from `.csv` file:

```{r}
# input data
df <- read.csv(here::here('data/fake_sample.csv'))
df$type <- as.factor(df$type)
df$geno <- as.factor(df$geno)
```

```{r, echo=FALSE}
DT::datatable(df, options = list(scrollX = TRUE), class = "cell-border stripe hover compact")
```

Prepare input parameters for the `lifelihood()` function:

```{r}
# name of the columns of the clutchs into a single vector
clutchs <- c(
  "clutch_start1", "clutch_end1", "clutch_size1",
  "clutch_start2", "clutch_end2", "clutch_size2"
)

# seeds value (must of length 4!)
seeds <- c(1, 2, 3, 4)
```

<br>

# Get the results
***

## All default parameters

```{r}
# run the analysis
results <- lifelihood(
  df = df,
  path_config = "../config.yaml",
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "mor_start",
  death_end = "mor_end",
  covariates = c("geno", "type"),
  seeds = seeds,
  matclutch = FALSE,
  model_specs = c("gam", "lgn", "wei"),
)
summary(results)
```

<br>

## Customize parameter boundaries

The parameter boundaries are stored in `parameter_ranges`:

```{r}
bounds_df <- results$parameter_ranges
```

```{r, echo=FALSE}
DT::datatable(bounds_df, options = list(scrollX = TRUE), class = "cell-border stripe hover compact")
```

If you want to customize them for your analysis, just do:

```{r}
bounds_df[bounds_df$name == "survival_shape", "min"] <- 2.0e-02

bounds_df[bounds_df$name == "expt_death", "max"] <- 30.0
```

```{r, echo=FALSE}
DT::datatable(bounds_df, options = list(scrollX = TRUE), class = "cell-border stripe hover compact")
```

Once it's changed, you just have to call `lifelihood()` again with the `param_bounds_df` argument:

```{r}
results <- lifelihood(
  df = df,
  path_config = "../config.yaml",
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "mor_start",
  death_end = "mor_end",
  covariates = c("geno", "type"),
  seeds = seeds,
  matclutch = FALSE,
  model_specs = c("gam", "lgn", "wei"),
  param_bounds_df = bounds_df,
)
summary(results)
```

# Get specific results
***

Get vector of estimation:

```{r}
results$effects
```

Get maximum likelihood found:

```{r}
results$likelihood
```