---
title: "Create simulation inputs"
vignette: >
  %\VignetteIndexEntry{create-simulation-input}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---

`create_simulation_input()` lets you simulate life histories from parameter
values that you choose yourself. It is useful when you want to explore a model
before fitting it, create a known synthetic population, or understand how
covariate effects change simulated maturity, reproduction, and mortality.

```{mermaid}
flowchart LR
  data["data<br/>covariates + sex + counts"] --> input["create_simulation_input()"]
  effects["effects<br/>chosen parameter values"] --> input
  config["config<br/>which formulas are fitted"] --> input
  input --> results["lifelihoodResults-like object"]
  results --> pred["prediction()"]
  results --> sim["simulate_life_history()"]
```

## Load packages

```{r}
devtools::load_all()
library(tidyverse)
```

## Create the population to simulate

The `data` argument describes the individuals you want to simulate. It must
contain the covariate columns and the sex column. Here, each row is one
combination of factor levels, and `n_individuals` says how many individuals to
create for that combination.

```{r}
population <- crossing(
  par = as.factor(c("none", "low", "high")),
  spore = as.factor(c("absent", "present")),
  sex = 0
) |>
  mutate(n_individuals = 100)

population
```

## Define the fitted formulas

The config tells `lifelihood` which parameters exist and which covariates each
parameter uses. This example uses:

- `par` and `spore` for mortality
- `par` for maturity
- `par` and `spore` for reproduction timing
- `par` for the number of offspring

```{r}
simulation_config <- list(
  mortality = list(
    expt_death = 1,
    survival_param2 = 1,
    ratio_expt_death = "not_fitted",
    prob_death = "not_fitted",
    sex_ratio = "not_fitted"
  ),
  maturity = list(
    expt_maturity = 1,
    maturity_param2 = 1,
    ratio_expt_maturity = "not_fitted"
  ),
  reproduction = list(
    expt_reproduction = 1,
    reproduction_param2 = 1,
    n_offspring = 1,
    increase_death_hazard = 1,
    tof_decay = "not_fitted",
    increase_death_hazard_n_offspring = "not_fitted",
    lin_decrease_hazard = "not_fitted",
    quad_decrease_hazard = "not_fitted",
    lin_change_n_offspring = "not_fitted",
    quad_change_n_offspring = "not_fitted",
    tof_n_offspring = "not_fitted",
    fitness = "not_fitted"
  )
)
```

## Choose the effects

Effects are written on the lifelihood link scale. A negative value pushes the
response toward the lower bound, and a positive value pushes it toward the upper
bound.

For categorical covariates, give one effect for every non-reference level. In
this example, `par` has levels `none`, `low`, and `high`, so `par = c(...)`
contains two effects: one for `low` and one for `high`.

```{r}
effects <- list(
  expt_death = list(intercept = 0), #, par = c(0.5, -0.4), spore = -0.5),
  survival_param2 = 0,
  expt_maturity = list(intercept = 0), # par = c(-0.3, -0.6)),
  maturity_param2 = 0,
  increase_death_hazard = 0,
  expt_reproduction = list(intercept = 0), # par = c(0.3, 0.6), spore = -0.25),
  reproduction_param2 = 0,
  n_offspring = list(intercept = 0) #, par = c(0.2, 0.5))
)
```

Be careful that an intercept set at 0 on the Lifelihood scale corresponds to an intercept set at the mean between the min and max bounds of the parameter defined via `default_bounds_df()` or via the `param_bounds_df` argument in `create_simulation_input()`.

## Build the simulation input

`covariates` and `sex` are column names in `data`. The `n_per_combination`
argument names the column that should be used to expand each row into multiple
individuals.

If `param_bounds_df` is not supplied, `create_simulation_input()` uses
`default_bounds_df()` internally.

```{r}
pseudo_results <- create_simulation_input(
  effects = effects,
  data = population,
  covariates = c("par", "spore"),
  sex = "sex",
  config = simulation_config,
  dist = c("wei", "wei", "wei"),
  n_per_combination = "n_individuals"
)

bounds_df <- default_bounds_df(pseudo_results$lifelihoodData)
bounds_df$max[bounds_df$param == "expt_death"] <- 100
bounds_df$max[bounds_df$param == "survival_param2"] <- 2
bounds_df$max[bounds_df$param == "expt_maturity"] <- 10
bounds_df$max[bounds_df$param == "maturity_param2"] <- 2
bounds_df$max[bounds_df$param == "expt_reproduction"] <- 5
bounds_df$max[bounds_df$param == "reproduction_param2"] <- 2
bounds_df$max[bounds_df$param == "increase_death_hazard"] <- 0.1

pseudo_results$param_bounds_df <- bounds_df
pseudo_results$sample_size
```

> If your `data` already has one row per individual, omit `n_per_combination`.


## Verify values of the parameters on the response scale

Please check that the values provided on the Lifelihood scales (defined in `effects` above) are realistic on the response scale.

```{r}
prediction(pseudo_results, "expt_maturity", type = "response") |>
  as_tibble() |>
  bind_cols(pseudo_results$lifelihoodData$df) |>
  distinct(par, spore, value)
```

```{r}
prediction(pseudo_results, "maturity_param2", type = "response") |>
  as_tibble() |>
  bind_cols(pseudo_results$lifelihoodData$df) |>
  distinct(par, spore, value)
```

## Simulate life histories

The object returned by `create_simulation_input()` can be passed directly to
`simulate_life_history()`.

```{r}
visits <- data.frame(block = 1, visit = seq(0.001, 1000, by = 0.1))
pseudo_results$lifelihoodData$block <- "block"
pseudo_results$lifelihoodData$df$block <- 1

simulated <- simulate_life_history(
  pseudo_results,
  seed = 1,
  use_censoring = TRUE,
  visits = visits
)

simulated |> head()
```

## Refit the simulated data

The previous section generated one synthetic dataset from the chosen parameter
values. We can fit the same model to that dataset with `lifelihood()` and compare
the refitted estimates with the values used for simulation.

```{r}
simulation_config_path <- tempfile(fileext = ".yaml")
yaml::write_yaml(simulation_config, simulation_config_path)

simulated_for_fit <- simulated |>
  mutate(
    sex_start = 0,
    sex_end = pseudo_results$lifelihoodData$right_censoring_date
  )

max_n_clutches <- max(simulated$total_n_clutches, na.rm = TRUE)
clutchs <- generate_clutch_vector(max_n_clutches)

simulated_lifelihood_data <- as_lifelihoodData(
  df = simulated_for_fit,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "maturity_start",
  maturity_end = "maturity_end",
  clutchs = clutchs,
  # `simulate_life_history()` names the mortality event `mortality_start` /
  # `mortality_end`, so point the death columns at those.
  death_start = "mortality_start",
  death_end = "mortality_end",
  covariates = c("par", "spore"),
  dist = c("wei", "wei", "wei"),
  matclutch = FALSE
)

refit <- lifelihood(
  lifelihoodData = simulated_lifelihood_data,
  path_config = simulation_config_path,
  param_bounds_df = pseudo_results$param_bounds_df,
  raise_estimation_warning = FALSE,
  delete_temp_files = FALSE,
  n_fit = 3
)
```

Because this is one finite simulated sample, the refitted estimates are not
expected to be exactly equal to the values used for simulation.

```{r}
pseudo_results$effects |>
  select(parameter, name, expected = estimation) |>
  inner_join(
    refit$effects |>
      select(parameter, name, fitted = estimation),
    by = c("parameter", "name")
  ) |>
  mutate(relative_difference = (expected - fitted) / expected) |>
  arrange(parameter, name) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 3)))
```

```{r}
prediction(pseudo_results, "expt_reproduction", type = "response") |>
  as_tibble() |>
  bind_cols(pseudo_results$lifelihoodData$df) |>
  distinct(par, spore, value)

prediction(refit, "expt_reproduction", type = "response") |>
  as_tibble() |>
  bind_cols(refit$lifelihoodData$df) |>
  distinct(par, spore, value)
```

The fitted model also has a regular log-likelihood, AIC, and BIC:

```{r}
c(
  logLik = logLik(refit),
  AIC = AIC(refit),
  BIC = BIC(refit)
)
```

## Common patterns

- Use one row per individual when you want full control of the simulated sample.
- Use one row per factor-level combination plus `n_per_combination` when you
  want a balanced or deliberately unbalanced population design.
- Omit `param_bounds_df` for defaults from `default_bounds_df()`.
- Supply `param_bounds_df` when you want tighter or custom parameter ranges for
  a teaching example, sensitivity analysis, or simulation study.
