---
title: "Create simulation inputs"
vignette: >
  %\VignetteIndexEntry{create-simulation-input}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---

`create_simulation_input()` lets you simulate life histories from parameter
values that you choose yourself. It is useful when you want to explore a model
before fitting it, create a known synthetic population, or understand how
covariate effects change simulated maturity, reproduction, and mortality.

```{mermaid}
flowchart LR
  data["data<br/>covariates + sex + counts"] --> input["create_simulation_input()"]
  effects["effects<br/>chosen parameter values"] --> input
  config["config<br/>which formulas are fitted"] --> input
  input --> results["lifelihoodResults-like object"]
  results --> pred["prediction()"]
  results --> sim["simulate_life_history()"]
```

## Load packages

```{r}
library(lifelihood)
library(tidyverse)

theme_set(theme_minimal(base_size = 12))
```

## Create the population to simulate

The `data` argument describes the individuals you want to simulate. It must
contain the covariate columns and the sex column. Here, each row is one
combination of factor levels, and `n_individuals` says how many individuals to
create for that combination.

```{r}
population <- crossing(
  par = factor(c("none", "low", "high"), levels = c("none", "low", "high")),
  spore = factor(c("absent", "present"), levels = c("absent", "present")),
  sex = c(0, 1)
) |>
  mutate(
    n_individuals = case_when(
      sex == 1 ~ 25L,
      spore == "present" ~ 35L,
      TRUE ~ 30L
    ),
    sex_label = if_else(sex == 1, "male", "female")
  )

population
```

If your `data` already has one row per individual, omit `n_per_combination`.

## Define the fitted formulas

The config tells `lifelihood` which parameters exist and which covariates each
parameter uses. This example uses:

- `par` and `spore` for mortality
- `par` for maturity
- `par` and `spore` for reproduction timing
- `par` for the number of offspring

```{r}
simulation_config <- list(
  mortality = list(
    expt_death = "par + spore",
    survival_param2 = 1,
    ratio_expt_death = "not_fitted",
    prob_death = "not_fitted",
    sex_ratio = "not_fitted"
  ),
  maturity = list(
    expt_maturity = "par",
    maturity_param2 = 1,
    ratio_expt_maturity = "not_fitted"
  ),
  reproduction = list(
    expt_reproduction = "par + spore",
    reproduction_param2 = 1,
    n_offspring = "par",
    increase_death_hazard = "not_fitted",
    tof_decay = "not_fitted",
    increase_death_hazard_n_offspring = "not_fitted",
    lin_decrease_hazard = "not_fitted",
    quad_decrease_hazard = "not_fitted",
    lin_change_n_offspring = "not_fitted",
    quad_change_n_offspring = "not_fitted",
    tof_n_offspring = "not_fitted",
    fitness = "not_fitted"
  )
)
```

## Choose the effects

Effects are written on the lifelihood link scale. A negative value pushes the
response toward the lower bound, and a positive value pushes it toward the upper
bound.

For categorical covariates, give one effect for every non-reference level. In
this example, `par` has levels `none`, `low`, and `high`, so `par = c(...)`
contains two effects: one for `low` and one for `high`.

```{r}
effects <- list(
  expt_death = list(intercept = -2.3, par = c(0.5, -0.4), spore = -0.5),
  survival_param2 = -1.2,
  expt_maturity = list(intercept = -3.2, par = c(-0.3, -0.6)),
  maturity_param2 = -1.2,
  expt_reproduction = list(intercept = -3.5, par = c(0.3, 0.6), spore = -0.25),
  reproduction_param2 = -1,
  n_offspring = list(intercept = -1.5, par = c(0.2, 0.5))
)
```

## Build the simulation input

`covariates` and `sex` are column names in `data`. The `n_per_combination`
argument names the column that should be used to expand each row into multiple
individuals.

If `param_bounds_df` is not supplied, `create_simulation_input()` uses
`default_bounds_df()` internally.

```{r}
simulation_input <- create_simulation_input(
  effects = effects,
  data = population,
  covariates = c("par", "spore"),
  sex = "sex",
  config = simulation_config,
  model = "lgn",
  n_per_combination = "n_individuals"
)

simulation_input$sample_size
```

## Simulate life histories

The object returned by `create_simulation_input()` can be passed directly to
`simulate_life_history()`.

```{r}
simulated <- simulate_life_history(simulation_input, seed = 1)

simulated |>
  select(
    death_start,
    death_end,
    maturity_start,
    maturity_end,
    total_n_offspring
  ) |>
  head()
```

## Common patterns

- Use one row per individual when you want full control of the simulated sample.
- Use one row per factor-level combination plus `n_per_combination` when you
  want a balanced or deliberately unbalanced population design.
- Omit `param_bounds_df` for defaults from `default_bounds_df()`.
- Supply `param_bounds_df` when you want tighter or custom parameter ranges for
  a teaching example, sensitivity analysis, or simulation study.
