---
title: "Customize parameter boundaries and estimation warning"
vignette: >
  %\VignetteIndexEntry{customize-parameter-boundaries-and-estimation-warning}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


<div class="topVignette">If you haven't already check it, have a look at:</div>

- [What is the required data format to work with lifelihood?](required-data-format.html)
- [Setting up the configuration file](setting-up-the-configuration-file.html)
- [How to use the lifelihood package](how-to-use-lifelihood.html)

# Load libraries
***

```{r}
library(lifelihood)
library(tidyverse)
```

# Data preparation
***

Load the dataset from `.csv` file:

```{r}
# input data
df <- fakesample |>
  mutate(
    type = as.factor(type),
    geno = as.factor(geno)
  )
```

```{r}
df |> head()
```

Prepare input parameters for the `lifelihood()` function:

```{r}
# name of the columns of the clutchs into a single vector
clutchs <- c(
  "clutch_start1",
  "clutch_end1",
  "clutch_size1",
  "clutch_start2",
  "clutch_end2",
  "clutch_size2"
)
```

*Note: If you have a large number of clutches, it is easier to generate this vector programmatically, particularly if your dataset contains a large number of clutches. See the [Generate clutch names](generate-clutch-names.html) vignette.*

# Create the `lifelihoodData` object

```{r}
dataLFH <- lifelihoodData(
  df = df,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("geno", "type"),
  model_specs = c("gam", "lgn", "wei")
)
```

# Get the results
***

Let's run the analysis with default parameters.

```{r}
results <- lifelihood(
  lifelihoodData = dataLFH,
  path_config = get_config_path("config")
)
```

# Warning
***

## What is it?

When runnning `lifelihood()` function, you might encounter the following warning:

```r
## Warning in check_estimation(results_lifelihood = results): Estimation of
## 'increase_death_hazard' is close to the maximum bound:
## increase_death_hazard≃9.9864992332278. Consider increasing maximum bound.
```

This warning indicates that the estimation of the `increase_death_hazard` parameter is close to the maximum bound.

This is not an error, but a signal that the estimation of the `increase_death_hazard` parameter is approaching its upper limit. During the optimization process, the algorithm may have been constrained by the current parameter boundaries, potentially affecting the accuracy of the estimation.

To address this warning, you can consider increasing the maximum bound of the `increase_death_hazard` parameter. This will give the optimization algorithm more flexibility to find the best estimate.

## Customize parameter boundaries

You can get the parameter boundaries with the `default_bounds_df()` function and by passing the `lifelihoodData` object:

```{r}
bounds_df <- default_bounds_df(dataLFH)
bounds_df
```

Since 10 seems to not be high enough, let's try with 80:

```{r}
bounds_df[bounds_df$name == "increase_death_hazard", "max"] <- 80
```


Once it's changed, you just have to call `lifelihood()` again with the `param_bounds_df` argument:

```{r}
results <- lifelihood(
  lifelihoodData = dataLFH,
  path_config = get_config_path("config"),
  param_bounds_df = bounds_df
)
```

Now we don't get any warning!
