---
title: "Fitting your first model in lifelihood"
vignette: >
  %\VignetteIndexEntry{fitting-your-first-model-in-lifelihood}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---



<div class="topVignette">If you haven't check it yet, have a look at:</div>

- [what is the required data format to work with lifelihood?](required-data-format.html)
- [setting up the configuration file](setting-up-the-configuration-file.html)

# Load libraries
***

```{r}
library(lifelihood)
library(tidyverse)
```

# Data preparation
***

Load the dataset from `.csv` file:

```{r}
# input data
df <- datapierrick |>
  as_tibble() |>
  mutate(par = as.factor(par), geno = as.factor(geno), spore = as.factor(spore))

df |> head()
```

Prepare arguments for the `as_lifelihoodData()` function:

```{r}
# name of the columns of the clutchs into a single vector
clutchs <- generate_clutch_vector(28)
```

*Note: If you have a large number of clutches, it is easier to generate this vector programmatically. See the [Generate clutch names](generate-clutch-names.html) vignette.*

# Create the `lifelihoodData` object

`as_lifelihoodData()` creates a `lifelihoodData` object, which is a list containing all the information needed to run the lifelihood program of a given dataset of individual life history.

This function mostly takes as input your dataset, your column names.

The `dist` argument is a data frame with exactly three columns: `mortality`, `maturity`, and `reproduction`. Each row specifies one model to fit. Each cell must contain `"wei"` (Weibull), `"exp"` (exponential), `"gam"` (gamma), or `"lgn"` (log-normal). Use a one-row data frame to fit a single model, as below.

```{r}
dataLFH <- as_lifelihoodData(
  df = df,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "death_start",
  death_end = "death_end",
  matclutch = FALSE,
  covariates = c("par", "geno"),
  dist = data.frame(mortality = "wei", maturity = "gam", reproduction = "lgn")
)
```

# Get the results
***

## All default parameters

Once you have created your `lifelihoodData` object with `as_lifelihoodData()`, you can call the `lifelihood()` function to run the lifelihood program.

It returns a `lifelihoodResults` object, which is a list containing all the results of the analysis.

Here it's a minimalist usage of the function, where we only specify the `lifelihoodData` object, the configuration and the seeds to use (four integers used by Mersenne Twister pseudorandom number generator of the lifelihood program). The `raise_estimation_warning` argument will be the focus of the [next vignette](4-custom-param-boundaries-and-estimation-warning.html).

```{r}
config <- list(
  mortality = list(expt_death = "par", survival_param2 = 1, ratio_expt_death = 1),
  maturity = list(expt_maturity = 1, maturity_param2 = 1),
  reproduction = list(
    expt_reproduction = 1,
    reproduction_param2 = 1,
    n_offspring = 1
  )
)

results <- lifelihood(
  lifelihoodData = dataLFH,
  config = config,
  seeds = c(1, 2, 3, 4),
  raise_estimation_warning = FALSE
)
summary(results)
```

# Get specific results
***

The `lifelihoodResults` object is a list containing all the results of the analysis. We can get specific results by calling the list element.

```{r}
coef(results)
coeff(results, "expt_death")
coeff(results, "survival_param2")

AIC(results)
BIC(results)

logLik(results)

prediction(results, parameter_name = "expt_death") |> head()
prediction(results, parameter_name = "expt_death", type = "response") |> head()
```

# Fit several models at once
***

To compare distribution families, supply one row per model. For example, the
following fits a Weibull/gamma/log-normal model and an exponential model:

```{r}
data_multiple <- dataLFH
data_multiple$dist <- data.frame(
  mortality = c("wei", "exp"),
  maturity = c("gam", "exp"),
  reproduction = c("lgn", "exp")
)

results_multiple <- lifelihood(
  lifelihoodData = data_multiple,
  config = config,
  n_fit = 2,
  raise_estimation_warning = FALSE
)

comparison <- summary(results_multiple$all_models)
comparison
```

`n_fit` applies to every row of `dist`, so this example runs four fits. Each
replicate receives fresh random seeds. Leave `seeds = NULL` when `n_fit > 1`;
with `n_fit = 1`, explicit seeds are reused for each model.

All models share the configuration. In a batch with multiple rows, second
parameters (`survival_param2`, `maturity_param2`, and `reproduction_param2`) are
automatically set to `"not_fitted"` for exponential events, which have no second
parameter. Other events keep their configured formulas. For a single-model fit,
an exponential event with a fitted second parameter raises a warning and that
parameter is disabled before fitting.

`results_multiple` contains the fit with the highest log-likelihood. The
comparison table includes every model and replicate, sorted by AICc, with the
difference from the lowest AICc in its `ΔAICc` column. The first row of this table
can differ from the model with the highest log-likelihood because AICc accounts
for the number of parameters. Use the `fit` column to retrieve a particular fit:

```{r}
best_aicc <- results_multiple$all_models[[comparison$fit[1]]]
best_aicc$dist
```

Each fit stores only its own row of `dist`; prediction, simulation, and
goodness-of-fit therefore use that model's distributions. Batch fitting also
works with `group_by_group = TRUE`. When default parameter boundaries are used,
they are computed separately for each model. A supplied `param_bounds_df`
applies to every model in the batch.

# Next step
***

Now that you have seen how to use the package, you can go further and [customise your parameter boundaries and deal with estimation warnings](customize-parameter-boundaries-and-estimation-warning.html).
