---
title: "Fit interaction model using `group_by_group = TRUE`"
vignette: >
  %\VignetteIndexEntry{fitting-interaction-model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
code-annotations: hover
execute: 
  cache: false
embed-resources: true
---


## Create a `lifelihoodData` object

```{r}
#| label: setup
library(lifelihood)
library(tidyverse)

df <- datapierrick |>
  as_tibble() |>
  mutate(
    par = as.factor(par),
    geno = as.factor(geno),
    spore = as.factor(spore)
  )

clutchs <- generate_clutch_vector(28)

lifelihoodData <- as_lifelihoodData(
  df = df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "geno"),
  dist = c("wei", "gam", "lgn")
)

set.seed(42)
```

## Difference between default and interaction models

- Default model

```{r}
time_default <- system.time({
  results_default <- lifelihood(
    lifelihoodData,
    path_config = use_test_config("config_gbg"),
    group_by_group = FALSE # Default value
  )
})
time_default
```

- Interaction model using the `group_by_group` argument (default to `FALSE`)

```{r}
time_gbg <- system.time({
  results_gbg <- lifelihood(
    lifelihoodData,
    path_config = use_test_config("config_gbg"),
    group_by_group = TRUE
  )
})
time_gbg
```

Fitting interaction model with group by group is faster than default model.

- Comparison

```{r}
results_default$likelihood
results_gbg$likelihood
```

Log-likelihood is also higher with group by group indicating better convergence towards the maximum log-likelihood value.
