---
title: "Fitting fitness"
vignette: >
  %\VignetteIndexEntry{fitting-fitness}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
code-annotations: hover
execute: 
  cache: false
embed-resources: true
---

```{r}
#| label: setup
devtools::load_all()
library(tidyverse)

clutchs <- generate_clutch_vector(6)

lifelihoodData <- as_lifelihoodData(
  df = datalenski,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "death_start",
  death_end = "death_end",
  covariates = "Group",
  dist = c(mortality = "wei", maturity = "lgn", reproduction = "lgn")
)

config_fitness <- list(
  mortality = list(expt_death = 1, survival_param2 = 1),
  maturity = list(expt_maturity = 1, maturity_param2 = 1),
  reproduction = list(
    expt_reproduction = 1,
    reproduction_param2 = 1,
    fitness = 1
  )
)
```


## Fit lifetime reproductive success

To fit lifetime reproductive success, we need to replace the `n_offspring` parameter by the `fitness` parameter. 

We recommend using MCMC sampling to compute fitness confidence intervals instead of `se.fit` which is likely to give wrong results.

```{r}
results <- lifelihood(
  lifelihoodData,
  config = config_fitness,
  n_fit = 30
)
```

## Results

We can then predict fitness and its confidence interval:

```{r}
summary(results)

prediction(results, "fitness", type = "response")
```

## Using simulated data


```{r}
population <- crossing(group = as.factor(0), sex = 0) |>
  mutate(n_individuals = 100)

effects <- list(
  expt_death = list(intercept = 0),
  survival_param2 = 0,
  expt_maturity = list(intercept = 0),
  maturity_param2 = 0,
  expt_reproduction = list(intercept = 0),
  reproduction_param2 = 0,
  n_offspring = list(intercept = 0)
)

config <- list(
  mortality = list(expt_death = 1, survival_param2 = 1),
  maturity = list(expt_maturity = 1, maturity_param2 = 1),
  reproduction = list(
    expt_reproduction = 1,
    reproduction_param2 = 1,
    n_offspring = 1
  )
)

pseudo_results <- create_simulation_input(
  effects = effects,
  data = population,
  covariates = c("group"),
  sex = "sex",
  config = config,
  dist = c(mortality = "exp", maturity = "exp", reproduction = "exp"),
  n_per_combination = "n_individuals"
)

bounds_df <- default_bounds_df(pseudo_results$lifelihoodData)
bounds_df$max[bounds_df$param == "expt_death"] <- 110
bounds_df$min[bounds_df$param == "expt_death"] <- 110
bounds_df$max[bounds_df$param == "survival_param2"] <- 2
bounds_df$max[bounds_df$param == "expt_maturity"] <- 10
bounds_df$min[bounds_df$param == "expt_maturity"] <- 10
bounds_df$max[bounds_df$param == "maturity_param2"] <- 2
bounds_df$max[bounds_df$param == "expt_reproduction"] <- 5
bounds_df$min[bounds_df$param == "expt_reproduction"] <- 5
bounds_df$max[bounds_df$param == "reproduction_param2"] <- 2
bounds_df$max[bounds_df$param == "n_offspring"] <- 10
bounds_df$min[bounds_df$param == "n_offspring"] <- 10

pseudo_results$param_bounds_df <- bounds_df

visits <- data.frame(block = 1, visit = seq(0, 1000, by = 0.1))
pseudo_results$lifelihoodData$block <- "block"
pseudo_results$lifelihoodData$df$block <- 1
simulated_df <- simulate_life_history(pseudo_results, visits = visits)
```

## Refit

```{r}
max_n_clutches <- max(simulated_df$total_n_clutches, na.rm = TRUE)
clutchs <- generate_clutch_vector(max_n_clutches)

lifelihoodData <- as_lifelihoodData(
  df = simulated_df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "maturity_start",
  maturity_end = "maturity_end",
  clutchs = clutchs,
  death_start = "mortality_start",
  death_end = "mortality_end",
  covariates = "group",
  dist = c(mortality = "exp", maturity = "exp", reproduction = "exp")
)

config_fitness <- list(
  mortality = list(expt_death = 1),
  maturity = list(expt_maturity = 1),
  reproduction = list(expt_reproduction = 1, fitness = 1)
)
config_fitness$reproduction$fitness

default_bounds_df(lifelihoodData)
results <- lifelihood(
  lifelihoodData,
  config = config_fitness,
  n_fit = 10,
  delete_temp_files = FALSE
)

summary(results)

prediction(
  results,
  c("expt_death", "expt_maturity", "expt_reproduction", "fitness"),
  type = "response"
)
```