---
title: "generate-clutch-names2"
vignette: >
  %\VignetteIndexEntry{generate-clutch-names2}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


The `lifelihood()` function requires a `clutch` argument. It's a vector containing the names of the clutch columns. The order should be: first clutch first date, first clutch second date, first clutch clutch size, second clutch first date, first clutch second date, second clutch clutch size, and so on. If the observation with the most clutches is, for example, 10, then the vector must be of size 10 x 3 = 30 (3 elements per clutch: first date, second date and size).

For practical reasons, it is easier to generate this vector programmatically, particularly if your dataset contains a large number of clutches.

# Load libraries & data
***

```{r}
library(lifelihood)

df <- fakesample
df$type <- as.factor(df$type)
df$geno <- as.factor(df$geno)
```

```{r}
#| echo=FALSE

DT::datatable(
  df,
  options = list(scrollX = TRUE),
  class = "cell-border stripe hover compact"
)
```

# Generate clutch names

Let's say my clutch names looks like this: `clutch_start_1`, `clutch_end_1`, `clutch_size_1`, `clutch_start_2`, `clutch_end_2`, `clutch_size_2`, and so on. This goes up to `clutch_start_28`, `clutch_end_28`, `clutch_size_28`.

We can generate this vector with the following function:

```{r}
generate_clutch_vector <- function(N) {
  return(paste(
    "clutch",
    rep(c("start", "end", "size"), N),
    rep(1:N, each = 3),
    sep = "_"
  ))
}
clutchs <- generate_clutch_vector(28)
clutchs
```

The only disadvantage of this technique is that the clutch names have to have a specific pattern.
