---
title: "How to use lifelihood"
vignette: >
  %\VignetteIndexEntry{goodness-of-fit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---


## Load libraries

```{r}
library(lifelihood)
library(tidyverse)
```

## Fit a simple model

```{r}
df <- datapierrick |>
  as_tibble() |>
  mutate(
    par = as.factor(par),
    geno = as.factor(geno),
    spore = as.factor(spore)
  ) |>
  sample_n(120)

generate_clutch_vector <- function(N) {
  return(paste(
    "pon",
    rep(c("start", "end", "size"), N),
    rep(1:N, each = 3),
    sep = "_"
  ))
}

lifelihoodData <- as_lifelihoodData(
  df = df,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = generate_clutch_vector(28),
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "spore"),
  model_specs = c("wei", "gam", "exp")
)

results <- lifelihood(
  lifelihoodData,
  path_config = use_test_config("config_pierrick"),
  raise_estimation_warning = FALSE
)
```

## Goodness of fit

The goodness of fit simulate datasets from a fitted model (`results)`, refit the model on each simulated dataset (`nsim`), and compare simulated log-likelihood values to the original fit.

```{r}
gof <- goodness_of_fit(results, nsim = 5)
```

The `goodness_of_fit()` function returns an instance of class `lifelihoodGOF`, with the following attributes:

```{r}
gof$original_loglik
gof$simulated_loglik
gof$n_success
gof$n_failed
gof$p_lower_or_equal
```

You can also read the `gof$fits` attribute for all underlying fits. Use `gof$fits[[1]]` for the first one, `gof$fits[[2]]` for the second, and so on.

## Visualization

You can use the `plot()` S3 method on the output of `goodness_of_fit()`:

```{r}
plot(gof)
```

We can see here that the simulated datasets, when fitted, have a less good log-likelihood compared to the original fit. This might suggest that the original fit isn't that great.