---
title: "Standard errors and MCMC"
vignette: >
  %\VignetteIndexEntry{prediction-with-standard-errors-and-mcmc}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
code-annotations: hover
execute: 
  cache: true
embed-resources: true
---

## Create a `lifelihoodData` object

```{r}
#| label: setup
library(lifelihood)
library(tidyverse)

df <- datapierrick |>
  as_tibble() |>
  mutate(
    par = as.factor(par),
    geno = as.factor(geno),
    spore = as.factor(spore),
    block = rep(1:2, each = nrow(datapierrick) / 2)
  )

clutchs <- generate_clutch_vector(28)

lifelihoodData <- as_lifelihoodData(
  df = df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  block = "block",
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "spore"),
  dist = c("wei", "gam", "lgn")
)
```

## Standard errors

By default, lifelihood will not try to fit standard errors. But, you can use the `se.fit` argument for this purpose:

```{r}

## Fail to compute standard errors due to absence of convergence to the ML optimum
results_wrong <- lifelihood(
  lifelihoodData = lifelihoodData,
  path_config = use_test_config("example_config_se"),
  se.fit = TRUE,
  seed = c(103, 349, 1213, 1283)
)

## Change sees to converge to the ML optimum
set.seed(123)
results <- lifelihood(
  lifelihoodData = lifelihoodData,
  path_config = use_test_config("example_config_se"),
  se.fit = TRUE,
  n_fit = 5
)

## New model has better convergence
logLik(results_wrong)
logLik(results)

summary(results)
```

Now if we have a look at the estimations we have standard errors:

```{r}
results$effects |> as_tibble()
```

## Prediction

We can predict with standard errors.

- Default scale

```{r}
prediction(results, "expt_death", se.fit = TRUE) |>
  as_tibble() |>
  sample_n(5)
```

- Response scale

```{r}
prediction(results, "expt_death", type = "response", se.fit = TRUE) |>
  as_tibble() |>
  sample_n(5)
```


## SE fails with interaction model with MCMC

> MCMC stands for **M**arkov **C**hain **M**onte **C**arlo.

### Fitting with MCMC

```{r}
results <- lifelihood(
  lifelihoodData = lifelihoodData,
  path_config = use_test_config("example_config_mcmc"),
  MCMC = 30
)
```

### Visualization

We can represent the confidence interval computed thanks to the standard errors:

```{r}
plot_fitted_event_rate(
  results,
  interval_width = 10,
  event = "mortality",
  use_facet = TRUE,
  groupby = "par",
  xlab = "Age (days)",
  ylab = "Fitted Mortality Rate",
  se.fit = TRUE
)
```