---
title: "Prediction"
vignette: >
  %\VignetteIndexEntry{prediction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
code-annotations: hover
execute: 
  cache: true
embed-resources: true
---

`prediction()` converts the coefficients of a fitted `lifelihood` model into
parameter values for individuals or combinations of covariates. It can return
predictions:

- for the data used to fit the model or for a new dataset;
- on the internal lifelihood scale or on the response scale;
- with standard errors or MCMC uncertainty when these were requested during
  model fitting.

This vignette covers the first two points and explains how predictions differ
between females and males. See the
[standard errors and MCMC vignette](prediction-with-standard-errors-and-mcmc.html)
for uncertainty estimates.

## Fit an example model

We first prepare `datapierrick` and fit the model used in the development
examples. The mortality formula in `config_pierrick` models `expt_death` as a
function of `par` and also fits `ratio_expt_death`, which represents the
male-to-female ratio in expected longevity.

```{r}
#| label: setup
#| output: false
library(lifelihood)
library(tidyverse)

df <- datapierrick |>
  as_tibble() |>
  mutate(par = as.factor(par), spore = as.factor(spore))

generate_clutch_vector <- function(n) {
  paste(
    "clutch",
    rep(c("start", "end", "size"), n),
    rep(seq_len(n), each = 3),
    sep = "_"
  )
}

lifelihood_data <- as_lifelihoodData(
  df = df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = generate_clutch_vector(28),
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "spore"),
  dist = c("wei", "gam", "lgn")
)

results <- lifelihood(
  lifelihoodData = lifelihood_data,
  path_config = use_test_config("config_pierrick"),
  seeds = c(3699, 783, 5401, 6502),
  raise_estimation_warning = FALSE
)
```

`parameter_name` must identify a parameter fitted by the model. The available
parameters can be read from the results:

```{r}
results$effects |> distinct(event, parameter)
```

## Predict for the fitted data

When `newdata` is omitted, `prediction()` returns one value for each individual
in the data used to fit the model.

```{r}
expected_longevity <- prediction(
  results,
  parameter_name = "expt_death",
  type = "response"
)

length(expected_longevity)
head(expected_longevity)
```

The result is a numeric vector when neither standard errors nor MCMC
predictions are requested. It can therefore be added directly to the original
data:

```{r}
df |>
  select(sex, par, spore) |>
  mutate(expected_longevity = expected_longevity) |>
  head()
```

The same approach applies to any fitted parameter:

```{r}
prediction(results, parameter_name = "expt_reproduction", type = "response") |>
  head()

prediction(
  results,
  parameter_name = "reproduction_param2",
  type = "response"
) |>
  head()
```

## Choose the prediction scale

By default, `type = "link"`. This returns the linear predictor on the internal
lifelihood scale:

```{r}
prediction(results, parameter_name = "expt_death", type = "link") |>
  head()
```

Use `type = "response"` for values on the parameter's original scale. For
`expt_death`, these values are expected longevity in the time unit used by the
input data:

```{r}
prediction(results, parameter_name = "expt_death", type = "response") |>
  head()
```

When a male ratio parameter is fitted, sex-specific values are applied only on
the response scale. The link-scale prediction therefore emits a message when
the prediction data include males.

## Predict on new data

Pass a data frame to `newdata` to predict for selected combinations of
covariates. It must contain:

- the sex column defined in `as_lifelihoodData()`;
- every covariate in the formula of the requested parameter.

Factor covariates must use levels from the fitted data. Preserving all original
levels is useful even when only a subset is present in `newdata`.

```{r}
newdata <- crossing(par = levels(df$par), sex = c(0, 1)) |>
  mutate(
    par = factor(par, levels = levels(df$par)),
    sex_label = if_else(sex == 0, "female", "male")
  )

newdata |>
  mutate(
    expected_longevity = prediction(
      results,
      parameter_name = "expt_death",
      newdata = newdata,
      type = "response"
    )
  )
```

Only `par` is required here because the fitted formula for `expt_death` is
`par`. Other columns from the original dataset are not needed.

## Predictions for females and males

For `expt_death` and `expt_maturity`, lifelihood can fit a corresponding ratio
parameter:

- `ratio_expt_death`;
- `ratio_expt_maturity`.

On the response scale, the prediction for a male is the female prediction
multiplied by the fitted ratio. The previous table therefore has different
values for females and males even when they have the same `par` level.

The ratio itself can also be predicted:

```{r}
newdata |>
  mutate(
    longevity_ratio = prediction(
      results,
      parameter_name = "ratio_expt_death",
      newdata = newdata,
      type = "response"
    )
  )
```

The ratio is `NA` for females because it is only used to modify male
predictions. If the corresponding ratio parameter is not fitted, no additional
male adjustment is applied.

## Prediction options

The return value changes when uncertainty is requested:

| Arguments | Return value |
|---|---|
| Default | Numeric vector of fitted values |
| `se.fit = TRUE` | Data frame with `fitted` and `se.fitted` |
| `mcmc.fit = TRUE` | Data frame with `fitted`, `mcmc_est`, and `mcmc_se` |
| `mcmc.fit = TRUE, keep_mcmc_samples = TRUE` | List containing `pred` and `mcmc_samples` |

`se.fit = TRUE` requires a model fitted with `lifelihood(..., se.fit = TRUE)`.
Similarly, `mcmc.fit = TRUE` requires a model fitted with
`lifelihood(..., MCMC = n)`, where `n` is greater than zero.

For example, after fitting with MCMC:

```{r}
#| eval: false
mcmc_prediction <- prediction(
  results,
  parameter_name = "expt_death",
  newdata = newdata,
  type = "response",
  mcmc.fit = TRUE,
  keep_mcmc_samples = TRUE
)

mcmc_prediction$pred
mcmc_prediction$mcmc_samples
```
