---
title: "Required data format"
vignette: >
  %\VignetteIndexEntry{required-data-format}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#library(lifelihood)
devtools::load_all()
```

# Introduction
***

As life history data is by nature heterogeneous between observations from the same sample, the format of the data required can be somewhat tricky. Under the hood, `lifelihood` creates a text file containing the data (and other parameters), and then use this file for the (core) source code. Fortunately, `lifelihood` automatically transform the dataframe into the text file with the right format.

In this vignette, we'll look at how the dataframe containing your data needs to be formatted for the Lifelihood programme to work properly.

# Data preparation
***

Let's create a simple dataset with just 7 observations and the following columns:

- `sex` Column name containing the sex of the observations.
- `sex_start` Column name containing the first date of the interval in which the sex was determined.
- `sex_end` Column name containing the second date of the interval in which the sex was determined.
- `maturity_start` Column name containing the first date of the interval in which the maturity was determined.
- `maturity_end` Column name containing the second date of the interval in which the maturity was determined.
- `clutchs` Vector containing the names of the clutch columns. The order should be: first clutch first date, first clutch second date, first clutch clutch size, second clutch first date, first clutch second date, second clutch clutch size, and so on. If the observation with the most clutches is, for example, 10, then the vector must be of size 10 x 3 = 30 (3 elements per clutch: first date, second date and size).
- `death_start` Column name containing the first date of the interval in which the death was determined.
- `death_end` Column name containing the second date of the interval in which the death was determined.
- `geno` Column name of the first column to add in the input data file

```{r}
df <- data.frame(
  sex = c(0, 0, 0, 0, 0, 0, 0),
  sex_start = c(1, 3, 2, 10, 3, 4, 5),
  sex_end = c(2, 4, 3, 11, 4, 5, 6),
  maturity_start = c(2, 1, 0, 1, 0, 2, 1),
  maturity_end = c(4, 2, 1000, 2, 1000, 3, 2),
  clutch_start1 = c(3, 2, NA, 2, NA, 3, 2),
  clutch_end1 = c(4, 3, NA, 3, NA, 4, 3),
  clutch_size1 = c(4, 6, NA, 5, NA, 2, 30),
  clutch_start2 = c(5, NA, NA, 5, NA, 4, 3),
  clutch_end2 = c(6, NA, NA, 6, NA, 5, 4),
  clutch_size2 = c(5, NA, NA, 7, NA, 10, 5),
  clutch_start3 = c(7, NA, NA, 6, NA, NA, 5),
  clutch_end3 = c(8, NA, NA, 7, NA, NA, 6),
  clutch_size3 = c(1, NA, NA, 1, NA, NA, 2),
  death_start = c(8, 11, 0, 11, 0, 7, 9),
  death_end = c(12, 11, 1, 12, 1, 8, 10),
  geno = c(1, 3, 1, 0, 2, 0, 1)
)
```

```{r}
#| echo=FALSE
DT::datatable(
  df,
  options = list(scrollX = TRUE),
  class = "cell-border stripe hover compact"
)
```

As you can see, some observations made more ponts, leading to the presence of NULL values.

*One* row of the dataset should represent the **life history** of *one* observation.

# Next step
***

Once our dataframe has the right format, we now have to [create a configuration file](setting-up-the-configuration-file.html).
