---
title: "Simulation and visit masks"
vignette: >
  %\VignetteIndexEntry{visit-masks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---

`lifelihood` offers a `lifelihood::simulate_life_history()` function that lets you simulate new observations based on estimates you made.

## Fitting

First, we need to fit the model with `lifelihood::lifelihood()`:


```{r}
#| label: setup
library(lifelihood)
library(tidyverse)

df <- datapierrick |>
  mutate(
    par = as.factor(par),
    spore = as.factor(spore),
    block = rep(1:2, each = nrow(datapierrick) / 2)
  )

generate_clutch_vector <- function(N) {
  return(paste(
    "clutch",
    rep(c("start", "end", "size"), N),
    rep(1:N, each = 3),
    sep = "_"
  ))
}
clutchs <- generate_clutch_vector(28)

lifelihoodData <- as_lifelihoodData(
  df = df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  block = "block",
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "spore"),
  model_specs = c("wei", "lgn", "wei")
)

results <- lifelihood(
  lifelihoodData,
  path_config = use_test_config("config_pierrick")
)

summary(results)
```

## Default simulations

By default, `lifelihood` will simulate all life history events (maturity, reproduction, and death):

```{r}
simulate_life_history(results) |> head()
```

But you can specify which event you want:

```{r}
simulate_life_history(results, event = "maturity") |> head()
```

## Simulations with visit masks

`lifelihood` lets you specify visit masks that are used to simulate data that more closely reflects how the original data was measured by adding constraints to the interval dates.

If you scroll to the top, you'll see that we passed a column name from our dataframe to the `block` argument. This column represents the block to which each individual belongs. Use `get_visits()` to retrieve the visit masks inferred from the original dataset:

```{r}
visits <- get_visits(lifelihoodData)
visits |> head()
```

Then pass these visits explicitly with `use_censoring = TRUE` to include censoring time intervals in the simulation:

```{r}
results |>
  simulate_life_history(
    event = "maturity",
    use_censoring = TRUE,
    visits = visits
  ) |>
  head()
```

You can also provide your own visit masks with the `visits` argument. It must be a dataframe with 2 columns: `block` (the same name as passed to `lifelihood::as_lifelihoodData()` in the `block` argument) and exactly `visit`. For each block, `visit` corresponds to the ages at which the events of individuals were recorded.

Let's define a custom visits dataframe:

```{r}
custom_visits <- tibble(
  block = rep(1:2, nrow(datapierrick) / 2),
  visit = seq_len(nrow(datapierrick))
)
custom_visits |> head()
```

Now we can pass this to the `simulate_life_history()` function:

```{r}
results |>
  simulate_life_history(
    event = "maturity",
    use_censoring = TRUE,
    visits = custom_visits
  ) |>
  head()
```
