---
title: "Simulation and visit masks"
vignette: >
  %\VignetteIndexEntry{visit-masks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
embed-resources: true
---

`lifelihood` offers a `lifelihood::simulate_life_history()` function that lets you simulate new observations based on estimates you made.

## Fitting

First, we need to fit the model with `lifelihood::lifelihood()`:

```{r}
devtools::load_all()
```

```{r}
#| label: setup
#| cache: true
library(tidyverse)

df <- datapierrick |>
  mutate(
    par = as.factor(par),
    spore = as.factor(spore),
    block = rep(1:2, each = nrow(datapierrick) / 2)
  )

clutchs <- generate_clutch_vector(28)

lifelihoodData <- as_lifelihoodData(
  df = df,
  matclutch = FALSE,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  block = "block",
  death_start = "death_start",
  death_end = "death_end",
  covariates = c("par", "spore"),
  dist = c("wei", "lgn", "wei")
)

results <- lifelihood(
  lifelihoodData,
  path_config = use_test_config("config_pierrick")
)

summary(results)
```

## Default simulations

By default, `lifelihood` will simulate all life history events (maturity, reproduction, and death):

```{r}
simulate_life_history(results) |> head()
```

But you can specify which event you want:

```{r}
simulate_life_history(results, event = "maturity") |> head()
```

## Simulations with visit masks

`lifelihood` lets you specify visit masks that are used to simulate data that more closely reflects how the original data was measured by adding constraints to the interval dates.

If you scroll to the top, you'll see that we passed a column name from our dataframe to the `block` argument. This column represents the block to which each individual belongs. Use `get_visits()` to retrieve the visit masks inferred from the original dataset:

```{r}
visits <- get_visits(lifelihoodData)
visits |> head()
```

To use censoring intervals in the simulation, you can pass visits explicitly along with `use_censoring = TRUE`:

```{r}
results |>
  simulate_life_history(
    event = "maturity",
    use_censoring = TRUE,
    visits = visits
  ) |>
  head()
```

Ideally, we recommend to provide explicitly the date where each visit occured for each block in the simulation. Indeed, that we see that, by default, with the function `get_visits()` above, Lifelihood considers that there was no visits for block 1 between t=0 and t=7 (as no events occured in this time interval), also visits occured every day.

You can also provide your own visit masks with the `visits` argument. It must be a dataframe with 2 columns: `block` (the same name as passed to `lifelihood::as_lifelihoodData()` in the `block` argument) and exactly `visit`. For each block, `visit` corresponds to the ages at which the events of individuals were recorded.

Let's define a custom visits dataframe:

```{r}
custom_visits <- tibble(
  block = rep(1:2, nrow(datapierrick) / 2),
  visit = seq_len(nrow(datapierrick))
)
custom_visits |> head()
```

Now we can pass this to the `simulate_life_history()` function:

```{r}
results |>
  simulate_life_history(
    event = "maturity",
    use_censoring = TRUE,
    visits = custom_visits
  ) |>
  head()
```

## Details

Requesting `event = "reproduction"` also simulates maturity and mortality, because both events are required to determine when reproduction can occur (e.g., after the maturity and before the death). Visit masks apply to reproduction in both the standard simulation and the reproduction-survival trade-off simulation.

```{r}
results |>
  simulate_life_history(
    event = "reproduction",
    use_censoring = TRUE,
    visits = visits,
    seed = 1
  ) |>
  head()
```

The following rules define how visit masks are constructed and applied:

- Visit masks are block-specific. The `visits` data frame must contain the configured block column and a `visit` column, with visit times for every block represented in the simulation. When `newdata` is supplied, it must also contain the configured block column.
- `get_visits()` collects the observed sex, maturity, and clutch interval bounds. It removes missing values and the right-censoring date, then deduplicates and sorts visit times within each block.
- An event is censored by the last visit before it occured and the first visit after it occured. The start date is set to `NA` when the event precedes the first visit, and the end date is set to `NA` when it follows the last visit.
- Reproduction masks are computed from absolute clutch ages. Maturity is added to the first inter-clutch duration, later durations are accumulated, and clutches after mortality are removed before visit intervals are assigned.
- If several clutches from one individual occur between the same visits, they are represented by a single clutch slot and the values of clutch sizes are summed. The exact simulated age is kept and named `clutch_{i}`.
- The clutches are numbered chronologically.
- Reproduction values are `NA` for males. `total_n_offspring` is calculated after same-interval clutches have been merged, so aggregation does not change the lifetime total.
