---
title: "3 - How to use the lifelihood package"
author: "Nicolas RODE, Thomas LENORMAND"
date: "`r format(Sys.Date(), '%d-%B-%Y')`"
output: 
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
editor_options: 
  chunk_output_type: console
---

<div class="topVignette">If you haven't already check it, have a look at:</div>

- [what is the required data format to work with lifelihood?](1-required-data-format.html)
- [setting up the configuration file](2-setting-up-the-configuration-file.html)

# Load libraries
***

```{r}
# library(lifelihood)
devtools::load_all(compile = FALSE)
```

# Data preparation
***

Load the dataset from `.csv` file:

```{r}
# input data
df <- read.csv(here::here("data/fake_sample.csv"))
df$type <- as.factor(df$type)
df$geno <- as.factor(df$geno)
```

```{r}
#| echo=FALSE

library(kableExtra)
head(df) %>%
  kbl() %>%
  kable_styling() %>%
  kable_material_dark(c("striped", "hover"))
```

Prepare input parameters for the `lifelihoodData()` function:

```{r}
# name of the columns of the clutchs into a single vector
clutchs <- c(
  "clutch_start1", "clutch_end1", "clutch_size1",
  "clutch_start2", "clutch_end2", "clutch_size2"
)
```

# Create the `lifelihoodData` object

```{r}
data <- lifelihoodData(
  df = df,
  sex = "sex",
  sex_start = "sex_start",
  sex_end = "sex_end",
  maturity_start = "mat_start",
  maturity_end = "mat_end",
  clutchs = clutchs,
  death_start = "mor_start",
  death_end = "mor_end",
  covariates = c("geno", "type"),
  model_specs = c("gam", "lgn", "wei")
)
summary(data)
```

# Get the results
***

## All default parameters

```{r}
results <- lifelihood(
  lifelihoodData = data,
  path_config = here::here("config.yaml"),
  seeds = c(1, 2, 3, 4)
)
summary(results)
```

# Get specific results
***

Get estimations:

```{r}
results$effects
```

Get maximum likelihood found:

```{r}
results$likelihood
```

# Next step
***

Now that you have seen how to use the package, you can go further and [customise your parameter boundaries and deal with estimation warnings](4-custom-param-boundaries-and-estimation-warning.html).
